---
title: "DDS 6306 : Case Study 2"
author: "Suchismita Moharana"
date: "11/24/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
library(curl)
library(ggthemes)
library(plotly)
library(viridisLite)
library(viridis)
library(gdata)
library(class)
library(caret)
library(e1071)
library(recipes)
```

# Introduction

This report is to analyse data for existing employees of a company to find the reason of employee turnovers. Also, we will talk about interesting trends that we see in employee data.

```{r}
#reading datafile and putting into dataframe

df_case_study2 <- read.csv(curl("https://raw.githubusercontent.com/BivinSadler/MDS-6306-Doing-Data-Science-Fall-2019/Master/Unit%2014%20and%2015%20Case%20Study%202/CaseStudy2-data.csv"))

df_case_study2_no_attr <- read.csv(curl("https://raw.githubusercontent.com/BivinSadler/MDS-6306-Doing-Data-Science-Fall-2019/Master/Unit%2014%20and%2015%20Case%20Study%202/CaseStudy2CompSet%20No%20Attrition.csv"))
df_case_study2_no_attr <- add_column(df_case_study2_no_attr,  Attrition = NA, .after = "Age")

df_case_study2_no_sal <- read.xls("/Users/suchismita/Documents/GitHub/Interim/CaseStudy2CompSet No Salary.xlsx", sheet = 1, header = TRUE)
df_case_study2_no_sal <- add_column(df_case_study2_no_sal,  MonthlyIncome = NA, .after = "MaritalStatus")

df_case_study2_whole <- rbind (df_case_study2, df_case_study2_no_attr, df_case_study2_no_sal)

df_case_study2_whole <- df_case_study2_whole %>% mutate(JobSatisfaction = replace(JobSatisfaction, JobSatisfaction == 1, "1_Low"))
df_case_study2_whole <- df_case_study2_whole %>% mutate(JobSatisfaction = replace(JobSatisfaction, JobSatisfaction == 2, "2_Medium"))
df_case_study2_whole <- df_case_study2_whole %>% mutate(JobSatisfaction = replace(JobSatisfaction, JobSatisfaction == 3, "3_High"))
df_case_study2_whole <- df_case_study2_whole %>% mutate(JobSatisfaction = replace(JobSatisfaction, JobSatisfaction == 4, "4_Very High"))

df_case_study2_whole$JobSatisfaction <- as.factor(df_case_study2_whole$JobSatisfaction)
```


### Below barplot is giving idea about the number of employees for each job title.

```{r}
# Plotting Job Role wise count
df_case_study2_whole %>% count(JobRole) -> df_jobrole_count 

#p = 
  ggplot(df_jobrole_count, aes (x = reorder(JobRole, n), y = n, fill = reorder(JobRole, n) )) + geom_bar(stat = "identity") + scale_fill_viridis (discrete = TRUE, direction = -1) + scale_color_manual(values=c("black", "white")) + theme(axis.text.x = element_text(angle=45,hjust=1, size=8)) + xlab("\nJob Title") + ylab ("Employee Count\n") + labs(fill = "Job Title") + ggtitle("Count summary for job title")

#ggplotly(p)
```

### Attrition percentage for different job titles. 

The below graph clearly states that turnver rate is high for `Sales Representatives`.

```{r}
df_case_study2_whole %>% ggplot(mapping = aes (x = JobRole, fill = Attrition)) + geom_bar(position = "fill") + ylab("Attrition Ratio\n")  + xlab("Job Title")  + theme(axis.text.x = element_text(angle=45,hjust=1,size=8)) + ggtitle ("Attrition percentage for each job title") 




#p <- plot_ly(df_jobrole_count, x = ~JobRole, y = ~n, type = 'bar', marker = list(color = c('rgba(204,204,204,1)', 'rgba(204,204,204,1)','rgba(204,204,204,1)','rgba(204,204,204,1)','rgba(204,204,204,1)','rgba(204,204,204,1)', 'rgba(204,204,204,1)','rgba(204,204,204,1)','rgba(222,45,38,0.8)'))) %>% layout(title = "Attrition percentage for different job titles", xaxis = list(title = "Job Title"), yaxis = list(title = "Attrition Ratio (Y:N)"))

#p

```


### Attrition value for different job title and monthly income. 

The below graph is indicating that attrition rate is high for lower income groups (especially for employees earning less than `5000` per month. As employees move to high income group, they tend to stay.  

```{r}
p = df_case_study2_whole %>% ggplot(mapping = aes (x = MonthlyIncome, y = TotalWorkingYears, shape = Attrition) ) + geom_point(aes(color = Attrition)) + ylab("Work experience (In years)\n") + xlab("\nMonthly Income") + ggtitle ("Monthly Income vs Work Experience for Attrition rate") + theme(axis.text.x = element_text(angle=0,size=8)) #+  facet_wrap(vars(Attrition)) + facet_grid(rows = vars(Attrition))

ggplotly(p) 

```

### Attrition Rate for employees working overtime

Working overtime plays an important role in employee turnover. As you can see from the below chart, more than 55% of the employees who fall under `Attrition = Yes` category worked overtime.

```{r}
df_case_study2_whole %>% ggplot(mapping = aes (x = Attrition , fill = OverTime )) + geom_bar(position = "fill") + ylab("Overtime Ratio\n") + xlab("\nAttrition?") + theme(axis.text.x = element_text(angle=0,size=8)) + ggtitle ("Attrition Rate for employees working Over Time")
```


### Attrition Rate for employees who travel more for work 

People who travel more for work tend to leave the company.

```{r}
df_case_study2_whole %>% ggplot(mapping = aes (x = factor(BusinessTravel,levels = c("Travel_Frequently", "Travel_Rarely", "Non-Travel")), fill = Attrition)) + geom_bar(position = "fill") + theme(axis.text.x = element_text(angle=0,size=8)) + ylab("Attrition Ratio (Y:N)") + xlab("Travel Frequency") + ggtitle ("Attrition Rate for employee's travel frequency")
```


### Attrition distribution for different income & different experience holders.

Below bar graph is an indicator of increasing monthly income with more experience. So, there is a positive co-relation. Employees with more work experience becomes Manager or Directors.

```{r}
df_case_study2_whole %>% ggplot(mapping = aes (x = TotalWorkingYears  , y = MonthlyIncome, color = JobRole) ) + geom_point() +  xlab("Work Experience (in years)") + ylab("Monthly Income") + ggtitle ("Work Experience vs Monthly Income for each job title")



df_case_study2_whole %>% ggplot(mapping = aes (x = TotalWorkingYears  , y = MonthlyIncome, color = Attrition) ) + geom_point() +  xlab("Work Experience (in years)") + ylab("Monthly Income") + ggtitle ("Work Experience vs Monthly Income for attrition")
```



### Monthly Income distribution for different job title

```{r}

df_case_study2_whole %>% ggplot(mapping = aes (x = JobRole, y = MonthlyIncome)) + geom_boxplot() + xlab("Job Title") + ylab("Monthly Income\n\n") + ggtitle ("Relationship between Job Title vs Monthly Income (Box Plot)") + theme(axis.text.x = element_text(angle=45,hjust=1, size=8)) 
```



### Impact of Job satisfaction for attrition rate

Job Satisfaction should be an ideal parameter for attrition rate. But as you can see from the below diagram, job satisfaction is not having a high impact on the attrition rate of this company. However, people with high job satisfcation value tend to stay back.


```{r}
#Attrition Rate for different Job Roles  --not very significant but ok
df_case_study2_whole %>% ggplot(mapping = aes (x = as.factor(JobSatisfaction), fill = Attrition)) + geom_bar(position = "fill") + theme(axis.text.x = element_text(angle=45,hjust=1,size=8)) + ylab("Attrition Ratio (Y:N)") + xlab("Job Satisfaction Level") + ggtitle ("Attrition Rate for employees having different job satisfaction value") 

#Attrition Rate for different Job Roles  --not very significant but ok
df_case_study2_whole %>% ggplot(mapping = aes (x = JobRole , fill = as.factor(JobSatisfaction) )) + geom_bar(position = "fill") + ylab("Percentage of Job Satisfaction") + xlab("Job Title") + theme(axis.text.x = element_text(angle=45,hjust=1,size=8)) + ggtitle ("Job Satisfaction ratio for Job role") + labs(fill = "Job Satisfaction") 

```


## Naïve Bayes Modeling 
```{r}
#model1 = naiveBayes(Attrition~. , data = df_case_study2_whole)
#df1 = data.frame(BusinessTravel = "Non_Travel", OverTime = "Yes")
#predict(model1, df1, type= "raw")

model2 = naiveBayes(Attrition~. , data = df_case_study2_whole)
df2 = data.frame(JobRole = "Sales Representative", OverTime = "Yes", BusinessTravel = "Travel_Frequently")
predict(model2, df2, type= "raw")
```


```{r}
#df_case_study2_MI = df_case_study2_whole %>% filter(!is.na(MonthlyIncome))

iterations = 100
masterAcc = matrix(nrow = iterations)
splitPerc = 0.8

for(j in 1:iterations)
{
  
  trainIndices = sample(1:dim(df_case_study2_whole)[1],round(splitPerc * dim(df_case_study2_whole)[1]))
  trainset = df_case_study2_whole[trainIndices,]
  testset = df_case_study2_whole[-trainIndices,]
  
  model = naiveBayes(trainset[,c(3,17,24)],trainset$Attrition)
  table(predict(model,testset[,c(3,17,24)]),testset$Attrition)
  CM = confusionMatrix(table(predict(model,testset[,c(3,17,24)]),testset$Attrition))
  masterAcc[j] = CM$overall[1]
}

MeanAcc = colMeans(masterAcc)
MeanAcc
```





## Conclusion

In my opinion, attrition is directly dependent on low income, working over time and more work related travel.

The posterior probability says, the attrition rate would be high for those employees who are in low income group (like Sales Representative) doing overtime and who Travel frequently.


We have a presentation of this report in https://youtu.be/c8dK0_q4g3I

